---
title:  "Day 3"
author:
- Sonja Hartnack
- Valerie Hungerbühler

date: "`r Sys.Date()`"

output:
  beamer_presentation: default
  slidy_presentation: default
  ioslides_presentation: default
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)

set.seed(1104)
```



## Which values for a prior beta distribution?

- Based on expert opinion
 
    * Betabuster https://shiny.vet.unimelb.edu.au/epi/beta.buster/
    * R package PriorGen
'Based on the available literature the mean value for the sensitivity of a test is expected to be 0.90 and we can be 95% sure that it is higher than 0.80.'

```{r, eval=FALSE, echo = TRUE}
library(PriorGen)
findbeta(themean=0.90, percentile=0.95,
         lower.v=FALSE, percentile.value=0.80)
```
Beta(27.79, 3.09)


## Model selection: inclusion of conditional dependencies between se and sp of different tests
- Pragmatic approach:
    - Look at 95% credibility intervals and histograms of posterior covariances: do they include a zero?
    - Are the other posteriors affected when including a covariance?
    - If either se or sp equal 1 (is perfect), then it will always be conditionally independent of the se or sp of the other test(s)
- Analytical approach:
    - DIC: deviance information criterion (Spiegelhalter, 2002)


## Ex 8 Covariates 
- Explore the data set 'echinococcus.xlsx'
    PCR for either E. multilocularis or E. granulosus, ELISA for both, eggs found by arecoline purgation, Taenia co-infection, age and sex 
- Run classical 'risk factor analysis': is sex, Taenia co-infection or age a risk factor for echinococcus (PCR-prevalence, seroprevalence or purges)? Obtain p-values and ORs with confidence intervals. 


## Ex 9 Covariates 
- Prepare the data set in the correct format (dump, add ones) for BLCM
- Run a model for three tests (assume a very high sensitivity for arecoline purgation)
- Try different priors
- Evidence of conditional dependencies?
- obtain dics
- Run models with covariates included 
    - 'runjags_version_cestode_update.R','model.cestode.bug', 'model_without_cestode.bug'
- Is there evidence for a covariate effect on the prevalence?
- Compare your findings with Ex.8
 
## Ex 10 Group work

- Please have a look at the paper from Kostoulas et al. 'STARD-BLCM: Standards for the Reporting of Diagnostic accuracy studies that use Bayesian Latent Class Models'

- Please discuus in your group for either the Bronsvoort, the Clothier or the Lahuerta paper if they comply with the STARD-BLCM guidelines.

- Please prepare a brief ppt, summarising the paper and assessing compliance with STARD-BLCM
