---
title:  "Day 2"
author:
- Sonja Hartnack
- Valerie Hungerbühler

date: "`r Sys.Date()`"

output:
  beamer_presentation: default
  slidy_presentation: default
  ioslides_presentation: default
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)

set.seed(1104)
```

## Ex 3 from yesterday continued
- Example of a COVID-19 data set
1. how to prepare the data set in the correct format
     * create_data_cassaniti.R
2. how to describe the model
     * model_final.bug
3. how to run the model in JAGS with runjags
     * runjags.version.R
4. how to check convergence
5. how to analyse the data

## Exercises
- Ex.3
    - Can you re-run the exercises?
    - Assess what happens if you add other covariances?
    - How many could you add and still have "meaningful results"?
    - Try different priors
    - You might also look at the runjags reference manual if ou find other things ou would like to customize?
    - Could you also extract the information for a single chain?
    
- Ex. 4 (Bonus)
    - Could you expand the model with a fourth test with simulated data?
    
    
## Prior choice
- Posterior is proportional to likelihood and to prior
$P(\theta|data) \propto P(data|\theta)\cdot P(\theta)$ 
- For the prior we need to choose a distribution and the values.

## Priors for binomial data
- Likelihood
$L(\pi) = \binom{n}{x} \pi^x(1-\pi)^{n-x}$ for $\pi \in (0,1)$

- log likelihood-kernel: 
$l(\pi)= x log \pi + (n-x)log(1-\pi)$
- first derivative
$\frac {dl(\pi)}{d\pi}=\frac{x}{\pi}-\frac{n-x}{1-\pi}$
- setting to zero gives MLE
$\hat{\pi}_{ML} =\frac{x}{n}$

## Combining binomial likelihood with a beta prior
- Likelihood:
$P(data|\pi) = \binom{n}{x} \pi^x(1-\pi)^{n-x}$ 
$P(data|\pi) \propto \pi^x(1-\pi)^{n-x}$
- Beta prior: $Beta(a,b)$
$P(\pi) \propto \pi^{\alpha-1}(1-\pi)^{\beta-1}$
- Posterior: 
$P(\pi|data) \propto \pi^x(1-\pi)^{n-x}\pi^{\alpha-1}(1-\pi)^{\beta-1}$
$P(\pi|data) \propto \pi^{x+\alpha-1}(1-\pi)^{n-x+\beta-1}$
$P(\pi|data)=Beta(\pi|x+\alpha,n-x+\beta)$

A pragmatic approach to choosing a prior distribution is to select a member of a specific family of distributions such that the posterior distribution belongs to the same family (conjugate prior distribution).

## Which values for a prior beta distribution?
- 1. Based on a previous publication: e.g. if 11 out of 303 individuals tested positive:
$P(\pi|data)=Beta(\pi|x+\alpha,n-x+\beta)$
$P(\pi|data)=Beta(\pi|11+1,303-11+1)$
$P(\pi|data)=Beta(\pi|12,293)$

```{r}
curve(dbeta(x, 12, 293), from=0, to=1)
```

## Which values for a prior beta distribution?
- 1. Based on a previous publication: e.g. if 110 out of 3030 individuals positive:
$P(\pi|data)=Beta(\pi|111,2921)$

```{r}
curve(dbeta(x, 111,2921), from=0, to=1)
```


## Which values for a prior beta distribution?
2. Based on expert opinion
    * Betabuster https://shiny.vet.unimelb.edu.au/epi/beta.buster/
    * R package PriorGen
'Based on the available literature the mean value for the sensitivity of a test is expected to be 0.90 and we can be 95% sure that it is higher than 0.80.'

```{r, eval=FALSE, echo = TRUE}
library(PriorGen)
findbeta(themean=0.90, percentile=0.95,
         lower.v=FALSE, percentile.value=0.80)
```
Beta(27.79, 3.09)

## Which values for a prior distribution
- data for prior should be independent of data for likelihood
- do a sensitivity analysis


## Ex.5
- Have a look at the excel file 'CLIA_LFIA_ELISA_final.xlsx'
- Use the file 'runjags_multiple_populations.R' to run the models 'blcm_A.R','blcm_B.R','blcm_C.R'
- Describe the different models, in which respect do they differ?
- Do you have other suggestions for 'better' models?


## Model selection: inclusion of conditional dependencies between se and sp of different tests
- Pragmatic approach:
    - Look at 95% credibility intervals and histograms of posterior covariances: do they include a zero?
    - Are the other posteriors affected when including a covariance?
    - If either se or sp equal 1 (is perfect), then it will always be conditionally independent of the se or sp of the other test(s)
- Analytical approach:
    - DIC: deviance information criterion (Spiegelhalter, 2002)


## Ex. 6 check with a pragmatic approach which model is better (data from Ex.5)

## Model selection DIC

![](C:\Users\shartn\Documents\Cost Harmony\Cost WS 2021\Trial files\spiegelhalter.pdf)

## Model selection criteria
- Question: if there are several possible models, which one is 'better'?
   - for example for nested models in generalised linear models, one could use deviance (likelihood ratio)
   - an alternative for non-nested models AIC Akaike information criterion
$AIC = 2k - 2ln(\hat{L})$ 
with $k$ the number of parameters and $ln(\hat{L})$ the maximum likelihood estimate.
    - AIC does not work in models with ninformative prior information
    - related BIC Bayesian information criterion
$BIC = k ln(n) - 2ln(\hat{L})$
with $n$ the number of data points.
    - DIC: deviance information criterion
$D(\theta)= -2log(p(y|\theta)) + C$

with $\theta$ unknown parameters, $p(y|\theta)$the likelihood and $C$ a constant
    - WAIC might be an alternative?
    
## Deviance information criterion (Spiegelhalter 2022, 2012)
- measure of ‘effective number of parameters’ $p_D$
 $DIC= D(\bar{\theta}) + 2p_D$ with $\bar{\theta}$ the expectation of $\theta$

- some criticisms: 
    - $p_D$ is not invariant to reparametrization
    - lack of consistenc
    - not based on a proper predictive criterion
    - has a weak theorertical foundation
    
## Ex. 7
  - With the Cassaniti data set, try to obtain DIC values for model with different conditional dependencies between sensitivities. 
  - look at 'DIC.R' and 'runjags_version_deviance.R'
  - (you might have a look at the WAIC.R file)
  
## Covariates: Latent variable logistic regression (Lewis 2012)
- A binomial regression model with a logit link function between the latent true prevalence and covariates associated with disease occurrence can be definded as follows, for covariate pattern $i$
$Pr(Y_i= y_i|n_i)=\binom{n_i}{k_i}q_i^{y_i}(1-q_i)^{n_i-y_i}$
where $q_i=Se\pi_i+(1-Sp)(1-\pi_i)$ and $log(\frac{\pi_i}{1-\pi_i})=x_i^T\beta$
- When $Se=1$ and $Sp=1$ then the model
reverts to the classical logistic regression model

## Ex 8 Covariates 
- Explore the data set 'echinococcus.xlsx'
    PCR for either E. multilocularis or E. granulosus, ELISA for both, eggs found by arecoline purgation, Taenia co-infection, age and sex 
- Run classical 'risk factor analysis': is sex, Taenia co-infection or age a risk factor for echinococcus (PCR-prevalence, seroprevalence or purges)? Obtain p-values and ORs with confidence intervals. 


## Ex 9 Covariates 
- Prepare the data set in the correct format (dump, add ones) for BLCM
- Run a model for three tests (assume a very high sensitivity for arecoline purgation)
- Try different priors
- Evidence of conditional dependencies?
- obtain DICs
  'runjags.version.deviance.R'
- Run models with covariates included 
    - 'runjags_version_cestode_update.R','model.cestode.bug', 'model_without_cestode.bug'
- Is there evidence for a covariate effect on the prevalence?
- Compare your findings with Ex.8
